---
title: " R Tip of the Day"
subtitle: "broom 🧹 package"
author: "Mutoni Henriette Isimbi"
date: "12, May, 2026"
format:
  revealjs: 
    slide-number: true
    theme: serif
    chalkboard: 
      buttons: true
    preview-links: auto
editor_options: 
  chunk_output_type: console
title-slide-attributes:
  data-background-color: "#89CFF0"
  data-color: "#003366"
---

```{r}
library(tidyverse)
library(broom)
library(gapminder)
```

##  {auto-animate="true" background="#89CFF0"}

::: {style="margin-top: 200px; font-size: 2.5em; color: black;"}
An introduction to broom 🧹
:::

## What is broom? {auto-animate="true" background="#89CFF0"}

-   broom is a package that converts statistical model output into tidy data frames

- This makes results easier to:
    -   visualize
    -   combine
    -   export
-   Works well with:
      `dplyr`, `ggplot2` and `purrr`
      
## broom Functions {auto-animate="true" background="#89CFF0"}

The three main functions are:

| Function    | Purpose                   |
|-------------|---------------------------|
| tidy()    | creates tidy coefficient tables       |
| glance()  | summarizes model performance  |
| augment() | adds predictions and residuals |


## Why Use broom? {auto-animate="true" background="#89CFF0"}

Without `broom`:

```{r}
#| echo: true
#| code-line-numbers: "1|3" 
model <- lm(lifeExp ~ gdpPercap, data = gapminder)

summary(model)
```

-   Output is messy and difficult to manipulate

## tidy() {auto-animate="true" background="#89CFF0"}

```{r}
#| echo: true
#| code-line-numbers: "1|3" 
model <- lm(lifeExp ~ gdpPercap, data = gapminder)

tidy(model)
```

-   Each coefficient becomes a row

# tidy() {auto-animate="true" background="#89CFF0"}

```{r}
#| echo: true
#| code-line-numbers: "1|2|3"
tidy(model) |>
  select("term","estimate")|>
  filter(term=="gdpPercap")
```

-   Easy to filter and visualize

## glance() {auto-animate="true" background="#89CFF0"}

```{r}
#| echo: true
glance(model)
```

- Returns one row model summaries

## augment() {auto-animate="true" background="#89CFF0"}

```{r}
#| echo: true
augment(model)
```

- Adds  fitted values and  residuals back onto the original dataset.

## broom + ggplot2 {auto-animate="true" background="#89CFF0"}

```{r}
#| echo: true
augment(model) |>
  ggplot(aes(lifeExp, .fitted)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, 
              intercept = 0,
              color = "red"
  ) +
  labs(
    title = "Actual vs Predicted",
    x = "Actual Life Expectancy",
    y = "Predicted Life Expectancy"
  )
```


## Comparing models {auto-animate="true" background="#89CFF0"}


```{r}
#| echo: true
model_log <- lm(lifeExp ~ log(gdpPercap), data = gapminder)
augment(model_log) |>
  ggplot(aes(x = lifeExp, y = .fitted)) +
  geom_point(alpha = 0.5) +
  geom_abline(  slope = 1,  intercept = 0, color = "red",
    linewidth = 1
  ) +
  labs(
    title = "Actual vs Predicted Life Expectancy",
    x = "Actual Life Expectancy",
    y = "Predicted Life Expectancy"
  )
```


## Summary  {auto-animate="true" background="#89CFF0"}

::: incremental
-   `tidy()` creates tidy coefficient tables
-   `glance()` summarizes model performance
-   `augment()` adds predictions and residuals
:::


## Real World Uses {auto-animate="true" background="#89CFF0"}

-   Machine learning 
-   Statistical automation
-   Model comparison


## Resources {auto-animate="true" background="#89CFF0"}

::: {style="text-align: center; margin-top: 2em"}
-   [broom documentation](https://broom.tidymodels.org/)

:::


## Thank You {auto-animate="true" background="#89CFF0"}

::: {style="margin-top: 200px; font-size: 2em; text-align: center;"}
Questions?
:::
